### 2020-10-27 manuscript/manuscript.rmd
fix table1 wideth when converted to PDF, also fix several typos

```{r preliminaries}
  library( tidyverse )
  library( magrittr )
  library( viridis )
  library( knitr )
  options(digits=2)
  library( gridExtra )
  library( ggpubr )
  library( grid )
  library( gtable )
  library( ggtree )
  library( ape )

@@

Resolving the placement of the root in the animal tree of life has proved very challenging [@laumer2019revisiting]. By synthesizing past phylogenomic studies and performing new analyses, we find that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some such analyses. Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency (Fig. 4). The 10-fold difference in category number seen in the more complex CAT models that support Porifera-sister does not improve model adequacy according to cross-validation, though. This suggests that we shouldn't priviledge these narrow analysis conditions that recover Porifera-sister over the much broader range of conditions that recover Ctenophora-sister.

@@

#### Phylogenetic distribution of support
To investigate the distribution of phylogenetic signal of the animal-root position in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices. We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to the rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanimalia and the full matrix). For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model. The two different phylogenetic trees passed to IQ-TREE (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to the rest of animals. The numbers of genes and sites supporting each hypothesis were calculated from IQ-TREE output and Perl scripts from a previous study [@shen2017contentious]. By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| > 2) or weak (|ΔlnL| < 2) phylogenetic signal as done in a previous study [@smith2020phylogenetic].

@@

**Extended Data Table 1.** Table 1. Overview of data matrices used in this study.
```{r table_study_summary}
table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
opts <- options(knitr.kable.NA = "")
knitr::kable(table_study_summary)

>>>>

```{r preliminaries}
  library( tidyverse )
  library( magrittr )
  library( viridis )
  library( knitr )
  options(digits=2)
  library( gridExtra )
  library( ggpubr )
  library( grid )
  library( gtable )
  library( ggtree )
  library( ape )
  library(kableExtra)

@@

Resolving the placement of the root in the animal tree of life has proved very challenging [@laumer2019revisiting]. By synthesizing past phylogenomic studies and performing new analyses, we find that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some such analyses. Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency (Fig. 4). The 10-fold difference in category number seen in the more complex CAT models that support Porifera-sister does not improve model adequacy according to cross-validation, though. This suggests that we shouldn't privilege these narrow analysis conditions that recover Porifera-sister over the much broader range of conditions that recover Ctenophora-sister.

@@

#### Phylogenetic distribution of support
To investigate the distribution of phylogenetic signal of the animal-root position in data matrices, we considered three major data matrices from three studies that had different topology between ML and BI using CAT model in our reanalysis, including Philippe2008, Ryan2013_est, and Whelan2017_full data matrices. We examined two hypotheses: Ctenophora-sister (T1) and Porifera-sister (T2) to the rest of metazoans, under a panel of evolutionary models with different outgroup schemes (Choanimalia and the full matrix). For IQ-TREE analysis in each data matrix, site-wise likelihood scores were inferred for both hypotheses using IQ-TREE (option -g) with the LG+G4 model. The two different phylogenetic trees passed to IQ-TREE 1.6.12 (via -z) were the tree where Ctenophora-sister and a tree modified to have Porifera placed as the sister to the rest of animals. The numbers of genes and sites supporting each hypothesis were calculated from IQ-TREE output and Perl scripts from a previous study [@shen2017contentious]. By calculating gene-wise log-likelihood scores between T1 and T2 for every gene, we considered a gene with an absolute value of log-likelihood difference of two as a gene with strong (|ΔlnL| > 2) or weak (|ΔlnL| < 2) phylogenetic signal as done in a previous study [@smith2020phylogenetic].

@@

**Extended Data Table 1.** Overview of data matrices used in this study.
```{r table_study_summary}
table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
opts <- options(knitr.kable.NA = "")
knitr::kable(table_study_summary, digits = 3, format.args = list(big.mark = ",",
  scientific = FALSE)) %>%
  kable_styling(latex_options="scale_down")
